---
title: "PTCAS Observation Hours"
author:
- Karl Holub <karl.holub@nested-knowledge.com>
- Molly Watkins <mwatkins2@css.edu>
output:
  html_document:
    df_print: paged
---
```{r setup, include=FALSE, echo=FALSE}
library(readxl)
library(dplyr)
library(readxl)
library(ggplot2)
library(maptools)
library(maps)
library(nymph) # install via: devtools::install_git('https://github.com/holub008/nymph')

parse_coords <- function(coords) {
  # all schools are in US, so we safely assume N latitude, W longitude
  # anything else will NA
  # TODO the conditionals allow for some human error & could be removed once that is fixed
  matches <- regmatches(coords, regexec("([0-9]+\\.[0-9]+) ?° N,? ([0-9]+\\.[0-9]+)° W", coords))
  lapply(matches, function(x){list(Latitude=as.numeric(x[2]), Longitude=-as.numeric(x[3]))}) %>%
    bind_rows()
}

data <- read_excel('./observation_hours.xlsx')
colnames(data) <- make.names(colnames(data))

data <- data %>% filter(!is.na(GPS.Coordinates)) # a couple new programs that we choose not to include
# TODO the coords for 193 seem incorrect, we ignore until corrected
data <- data[-193,]
data <- data %>% cbind(parse_coords(data$GPS.Coordinates)) %>%
  select(Instituition, Category, Minimum.Hours, Recommended.Hours, Latitude, Longitude)

categories <- data.frame(
  id = 1:6,
  Requirement.Category = c(
    'PT hours are required--a licensed PT must verify with signed form uploaded or online via PTCAS',
    'PT hours are required--no verification by a physical therapist',
    'PT hours are not required but are highly recommended',
    'PT hours are not required but are considered',
    'PT hours are not required or considered',
    'Other'
  )
)

# TODO this drops a program which doesn't have a category currently
oh_data <- data %>% 
  inner_join(categories, by=c('Category' = 'id')) %>%
  filter(Longitude < -50 & Latitude > 20) # TODO filtering these non-Americas for now
```

# About
We are studying the question of how PT programs Observation hours are time that prospective PT school students

# Dataset
The dataset was sourced from the [PTCAS website](http://aptaapps.apta.org/ptcas/observationhours.aspx) (for requirements) & [Google Maps](https://google.com/maps) (for program locations). It has the following shape:
```{r}
dim(oh_data)
```

meaning there are 243 programs & 7 (really 6) descriptive columns. Previewing it:

```{r}
head(oh_data, n=10)
```

Outlining the non-trivial columns:

* Category: The type of the requirement the institution has for observation hours
* Minimum.Hours: The minimum number of observation hours required. May be missing if:
    * The program doesn't require hours
    * The program doesn't publish an explicit number, in favor of "case by case" consideration. Students should still expect to need to have some hours for acceptance.
* Recommended.Hours: Optional and potentially different from minimum hours
    * Some programs supply a range (e.g. "400-600"") or lower bounded range (e.g. "100+""). For this analysis, only the lower bound is used.
* Latitude & Longitude: Approximate (hand selected on map) location of the center of the institution's campus


# Summary Statistics

## Requirement Category
```{r}
oh_data %>%
  ggplot() +
    geom_bar(aes(as.factor(Category))) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Requirement Category') +
    ylab('Number of Programs')
```

where the categories are:

```{r}
categories
```

83% of programs require observation hours.

## Required Hours
```{r}
summary(oh_data$Minimum.Hours)
```

The distribution of hours across all programs looks like:

```{r}
oh_data %>%
  filter(!is.na(Minimum.Hours)) %>%
  ggplot() +
    geom_histogram(aes(Minimum.Hours), bins = 15)
```

Peeking at distributions conditioned on the requirement category (specifically 1 & 2, which do impose a requirement):

```{r}
oh_data %>%
  filter(!is.na(Minimum.Hours) & Category %in% c(1,2)) %>%
  ggplot() +
    geom_violin(aes(as.factor(Category), Minimum.Hours))
  
```

The distributions are similar, with modes at 40 & 100 hours.

# Incorporating Geography

## Visualizing

### Points
First, we view each program as a point on the US map, where color indicates the hours requirement:
```{r fig.width=20, fig.height=15}
us <- map_data('state')

gg <- oh_data %>%
  ggplot()  +
    geom_polygon(data = us, aes(x=long, y = lat, group = group), fill='grey', color='white') +
  geom_point(aes(x=Longitude, y=Latitude, color = Minimum.Hours), size=15)

gg
```

There is heavy concentration of programs in the Northeast US, so we zoom in:

```{r fig.width=15, fig.height=15}
gg + coord_cartesian(xlim = c(-79, -67), ylim = c(35, 47))
```

### Heatmap

TODO: This isn't very pretty or effective, despite probably being the most important graph!
```{r fig.width=15, fig.height=11}
oh_data %>%
  filter(!is.na(Minimum.Hours)) %>%
  mutate(
    Latitude = as.integer(Latitude),
    Longitude = as.integer(Longitude)
  ) %>%
ggplot() +
  geom_tile(aes(x = Longitude, y = Latitude, fill = Minimum.Hours)) +
  geom_polygon(data = us, aes(x=long, y = lat, group = group), alpha=0, color='black')
```

There certainly appears to be a trend of decreasing hour trends as Longitude increases (moving towards the East). There also appears to be pockets of high requirements in the "central" east, which might suggest a network analysis, or ranged interaction effects in latitude and longitude.

## Correlations

### Main Effect Lat/Long

#### Visualizing
To visualize main effect correlations, we look at marginal x/y plots of latitude and longitude. Smoothing splines are used to trend the data.

```{r message=FALSE}
oh_data %>%
  filter(!is.na(Minimum.Hours)) %>%
  ggplot(aes(Latitude, Minimum.Hours)) +
    geom_point() +
    geom_smooth()
```

Hours look pretty flat across Latitude, with perhaps a slight decrease moving North.

```{r message=FALSE}
oh_data %>%
  filter(!is.na(Minimum.Hours)) %>%
  ggplot(aes(Longitude, Minimum.Hours)) +
    geom_point() +
    geom_smooth()
```
There is a definite decrease in hours moving west to east, with a possible jump upwards at the coast.

#### Inference (Non-zero Effects)

To determine if the main effect correlations are non-zero, we perform a bootstrap on the correlation coefficients, building 95% intervals:

```{r}
bs <- bootstrap(oh_data %>% filter(!is.na(Minimum.Hours)), corr_lat = cor(Latitude, Minimum.Hours, method='spearman'), corr_long = cor(Longitude, Minimum.Hours, method='spearman'))

summary.bootstrap(bs)
```

Matching our visual intuition, the Longitude correlation interval certainly does not include 0 (i.e. is a non-zero effect), while Latitude is less convincing (interval does include a 0 correlation). For future analyses, it is extremely likely that these two effects interact (e.g. due to high requirements in the Northeast US).

### Spatial Auto Correlation
TODO



